---
title: "EDA for final project"
author: "Tiki Zhai"
date: "`r Sys.Date()`"
format:
  html:
    toc: true       
    toc-depth: 2  
    number-sections: true
---

# Introduction

This document summarizes the structure of the dataset, performs exploratory data analysis (EDA), and identifies missing or inconsistent data. The goal is to prepare a clean dataset for stroke risk modeling using a neural network.

## Set upˆ

```{r setup, include=FALSE}
library(tidyverse)
```

## Load the Raw dataset.

```{r}
stroke_raw <- read_csv(here::here("healthcare-dataset-stroke-data.csv"))
```

# Exploratory Data Analysis

## Missing data Analysis

### Make each column either numeric or factor to better assess the missing rows.

```{r}
# Initial type conversion (before handling special missing cases)
stroke_raw <- stroke_raw |>
  mutate(
    age = as.numeric(age),
    avg_glucose_level = as.numeric(avg_glucose_level),
    hypertension = as.factor(hypertension),
    heart_disease = as.factor(heart_disease),
    stroke = as.factor(stroke),
    gender = as.factor(gender),
    ever_married = as.factor(ever_married),
    work_type = as.factor(work_type),
    residence_type = as.factor(Residence_type),
    bmi = bmi,  # placeholder for now
    smoking_status = smoking_status  # placeholder for now
  )
```

### The bmi column is in character, which makes the NA values not detectable, similarly with smoking_status column, need to fix them.

```{r}
# Convert "N/A" and "Unknown" to NA and recast types
stroke_raw <- stroke_raw |>
  mutate(
    bmi = as.numeric(na_if(bmi, "N/A")),
    smoking_status = as.factor(na_if(smoking_status, "Unknown"))
  ) |> select(-Residence_type)
```

### Explore the count and the percentage of the missing data with each column

```{r}
# Count and percent missing
stroke_raw |> 
  summarise(across(everything(), ~sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") |>
  mutate(percent_missing = missing_count / nrow(stroke_raw) * 100) |>
  arrange(desc(missing_count))
```

As we can see, there are only two columns with missing data which are smoking_status and bmi.

## Summary statistics.

### Numeric variables, summary statistics and visualizations

#### Numeric variables's summary table showing Minimum, maximum, median, mean, etc.

```{r}
stroke_raw |>
  select(age, avg_glucose_level, bmi) |>
  summary()
```

#### Numeric variable's data visualization

```{r}
# Age
stroke_raw |>
  ggplot(aes(x = age)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Age Distribution", x = "Age", y = "Count")

# BMI
stroke_raw |>
  ggplot(aes(x = bmi)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white") +
  labs(title = "BMI Distribution", x = "BMI", y = "Count")

```

### Density plot grouped by the outcome variable

```{r}
# BMI vs Stroke
stroke_raw |>
  ggplot(aes(x = stroke, y = bmi, fill = stroke)) +
  geom_boxplot() +
  labs(title = "BMI by Stroke Status", x = "Stroke", y = "BMI")

# Glucose by Stroke
stroke_raw |>
  ggplot(aes(x = stroke, y = avg_glucose_level, fill = stroke)) +
  geom_boxplot() +
  labs(title = "Glucose Level by Stroke Status", x = "Stroke", y = "Glucose")
```

### Class imbalance of the outcome variable

```{r}
stroke_raw |>
  count(stroke) |>
  mutate(prop = round(n / sum(n), 3))
```

# Data interpolation

## Interpolate the "bmi" column

```{r}
median_bmi <- median(stroke_raw$bmi, na.rm = TRUE)

stroke_raw <- stroke_raw |>
  mutate(bmi = if_else(is.na(bmi), median_bmi, bmi))

```

Chose median interpolation because it is robust to outliers and works well for clinical data

## Interpolate the "Smoking status" column

```{r}
stroke_raw <- stroke_raw |>
  mutate(smoking_status = fct_explicit_na(smoking_status, na_level = "Missing"))
```

## Inspect the missing status again.

```{r}
stroke_raw |> 
  summarise(across(everything(), ~sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") |>
  mutate(percent_missing = missing_count / nrow(stroke_raw) * 100) |>
  arrange(desc(missing_count))
```

# Data dictionary of the dataframe

## 📘 Data Dictionary

| Variable | Type | Description |
|--------------|--------------|--------------------------------------------|
| age | Numeric | Age of the patient |
| hypertension | Factor | 0 = No hypertension, 1 = Has hypertension |
| heart_disease | Factor | 0 = No heart disease, 1 = Has heart disease |
| ever_married | Factor | Marital status ("Yes" or "No") |
| work_type | Factor | Type of work: "Private", "Self-employed", etc. |
| residence_type | Factor | "Urban" or "Rural" |
| avg_glucose_level | Numeric | Average glucose level in blood (mg/dL) |
| bmi | Numeric | Body Mass Index (imputed with median if missing) |
| smoking_status | Factor | "formerly smoked", "never smoked", "smokes", or "Missing" |
| gender | Factor | "Male", "Female", or "Other" |
| stroke | Factor | **Target variable**: 0 = No stroke, 1 = Had stroke |
